Lock-free Skiplist-based Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-05 paper/paper.tex
Hunt heaps, concurrent SkipLists

There are several different criteria which allows reasoning about the correctness of concurrent
data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
effect at a single instant in time at so-called linearization points. Linearizability is composable ---
any data structure composed of linearizable objects is also linearizable.
\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
executed in parallel to be equivalent to the result of some sequential ordering of the same
operations.
\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
after a period of quiescence, i.e. an interval without active operations has passed.

@@

Traditional data structures such as the Heap have fallen out of favor;
instead, SkipLists \cite{pugh1990skip,pugh1998concurrent} have become the focus
of modern concurrent priority queue research
\cite{shavit2000skiplist,sundell2003fast,herlihy2012art,linden2013skiplist,alistarhspraylist}.
SkipLists are both conceptual simple as well as simple to implement; they also exhibit
excellent disjoint-access properties, and do not require rebalancing due to their
reliance on randomization for expected $O(\log n)$ time insertion and deletion bounds.

@@

\subsection{Shavit \& Lotan}

\citeauthor{shavit2000skiplist} were the first to propose the use of SkipLists
for priority queues \cite{linden2013skiplist}. Their initial implementation
\cite{shavit2000skiplist} builds on \citeauthor{pugh1998concurrent}'s concurrent
SkipList idea \cite{pugh1998concurrent} which uses one lock per node per level.
\citeauthor{herlihy2012art} \cite{herlihy2012art} later succeeded in making this
priority queue lock-free (but not linearizable) and provide a Java implementation
which we will examine further in the following.

% TODO

\subsection{Sundell \& Tsigas}

\subsection{Linden \& Jonsson}

@@

\section{Performance Results}

In this section, we compare the performance of several different concurrent priority
queue implementations:

\begin{itemize}
\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
      However, as the original code was written for the MIPS architecture, we chose to
      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
      instead.
\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
      According to the authors, the commercial Noble library includes a more efficient
      version of this data structure that can also handle duplicate priority values.
\item \textbf{Linden} Original code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
      is available under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
      SkipList implementation.
\end{itemize}

In each test run, the examined priority queue was initially filled with $2^15$
elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
for a total of $10$ seconds, where all \lstinline|Insert|
operations within this context implicitly choose a key uniformly at random from
the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
and we report on the average throughput.

All benchmarks were compiled with \verb|-O3| on GCC 4.8.2. Evaluations took place
on an 80-core Intel Xeon machine.

% TODO: Figures, interpretation. Mention that Noble requires pointers -> new/delete.

\begin{comment}
Sections / rough structure:
* Basic concepts and definitions. Linearizability, sequ./quiescent consistency,
  lock-free, wait-free, disjoint-access parallelism (one of the papers has good
  summaries of these).
  Maybe atomic primitives such as CAS (but probably not).
* Priority queue definitions, semantics, usages. Mention inherent
  non-scalability through DeleteMin().
* Skiplist, heap definitions, semantics, usages.
* Follow development of current state of the art from Hunt Heap ~> Shavit ~>
  Tsigas ~> Linden.
* Maybe benchmarks on mars.

Additional notes:
* Sketch development: Heap, Fraser Skiplist (Details, mention why linearizable),
  Sundell/Tsigas, other Skiplist-based PQs
* Modern developments: SprayList, Wimmer et al. Relation to other relaxed structures such as k-FIFO.
\end{comment}

@@

% Bounded priority queues
% PQ's with non-available primitives
% Other relaxed data structures -> k-FIFO
% ... ?

@@

>>>>

There are several different criteria which allows reasoning about the correctness of concurrent
data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
effect at a single instant in time at so-called linearization points. Linearizability is composable ---
any data structure composed of linearizable objects is also linearizable.
\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
executed in parallel to be equivalent to the result of some sequential ordering of the same
operations.
\emph{Quiescently consistent} \cite{shavit1996diffracting}
guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
after a period of quiescence, i.e. an interval without active operations has passed.

@@

Traditional data structures such as the Heap have fallen out of favor;
instead, SkipLists \cite{pugh1990skip} have become the focus of modern concurrent
priority queue research
\cite{shavit2000skiplist,sundell2003fast,herlihy2012art,linden2013skiplist,alistarhspraylist}.
SkipLists are both conceptual simple as well as simple to implement; they also exhibit
excellent disjoint-access properties, and do not require rebalancing due to their
reliance on randomization for expected $O(\log n)$ time insertion and deletion bounds.

@@

\subsection{Lotan \& Shavit}

\subsection{Sundell \& Tsigas}
\subsection{Linden \& Jonsson}

@@

\begin{comment}
Sections / rough structure:
* Basic concepts and definitions. Linearizability, sequ./quiescent consistency,
  lock-free, wait-free, disjoint-access parallelism (one of the papers has good
  summaries of these).
  Maybe atomic primitives such as CAS (but probably not).
* Priority queue definitions, semantics, usages. Mention inherent
  non-scalability through DeleteMin().
* Skiplist, heap definitions, semantics, usages.
* Follow development of current state of the art from Hunt Heap ~> Shavit ~>
  Tsigas ~> Linden.
* Maybe benchmarks on mars.
\end{comment}

@@

@@

% TODO: Remove me.
\nocite{*}
